A Model for Extracting Time Delays from Noisy Time Series Data

### 2013-08-23 doc/Paper1/ms.tex
Random work

However, any inference of $H_0$ or other cosmological parameters
from the gravitational lensing technique inherits
the uncertainty from the measurements of the time delays themselves. A
thorough investigation of all known sources of uncertainty is an integral part
of measuring cosmological parameters, and thus the purpose of this paper is to
develop and present a reliable method for estimating the uncertainty on the
values of the time delays themselves. The method is Bayesian
\citep{2004kats.book.....O} so the results obtained are probability
distributions over the space of possible solutions for the time delays, and the uncertainties in the time delays
can be summarised by the widths of these probability distributions. To describe
the underlying time variation of the quasar and the effects of microlensing on
the light curves, we use Gaussian Processes (GPs), a family of probability
distributions over function space that has computationally convenient
properties \citep{rasmussen}.

@@

In Section~\ref{sec:bayes} we briefly discuss Bayesian Inference, and in
Section~\ref{sec:sampling} we describe our specific
modelling assumptions. The computational implementation is described in
Section~\ref{sec:computation}. In Section~\ref{sec:tests} we test the inference
with two simulated data sets, one ``easy'' data set and one ``hard'' data set.
In Section~\ref{sec:b1608} we apply the model to B1608+656.

@@

We begin by defining our model for the prior knowledge about how the time
delays give rise rise to the observed data. This will determine the sampling
distributions, and also the set of parameters (including nuisance parameters) to
be inferred. Our choice of prior distribution for the parameters will be
discussed in Section

\section{Sampling Distributions}\label{sec:sampling}
The intrinsic quasar variability, in magnitudes, and in whatever band the
observations are carried out in, can be described by a function $q(t)$. For a
two-image system, the resulting light curves, if observed without gaps and
noise, would be described by the functions
\begin{eqnarray}
y_1(t) &=& m_1 + f(t) + \mu_1(t) \\
y_2(t) &=& m_2 + f(t - \tau_2) + \mu_2(t) \\
\end{eqnarray}
where $\{m_1, m_2\}$ are the mean magnitudes of the images,
$\tau_2$ is the time delay of image 2 relative to image 1, and the
functions $\{\mu_1(t), \mu_2(t)\}$ describe any perturbations that are not
shared by both images. A major source of perturbations is the effect of
microlensing on each of the light curves. The data then consist of
noisy, irregularly sampled measurements of the light curves
$\{y_1(t), y_2(t)\}$.

@@

When we use Bayes' rule, the flags $\{k_1, k_2, ..., k_N\}$ and measurement
times $\{t_1, t_2, ..., t_N\}$ will be subsumed into the prior information. The
data will be the measurements
\begin{eqnarray}
\boldsymbol{D} = \{Y_1, Y_2, ..., Y_N\}.
\end{eqnarray}

The sampling distributions (and hence the likelihood function) will then be
written as a probability distribution for the measurements $\data$
given the time delay $\tau_2$ along with some nuisance parameters.

\begin{table*}
\begin{tabular}{lll}
\hline
{\bf Parameter} &	{\bf Meaning}			& {\bf Prior}\\
\hline
$\tau_2$	&	Time delay relative to image 1	& Cauchy$\left(0, 0.1\tau_{\rm range}\right)T\left(\tau_{\rm min},\tau_{\rm max}\right)$\\
$\{m_1, m_2\}$	&	Mean magnitude for each image	& Uniform$(m_{\rm min}, m_{\rm max})$\\
$L$		&	Length scale of QSO variations	& LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\beta$		&	Size of QSO variations		& LogUniform$(\beta_{\rm min}, \beta_{\rm max})$\\
$A$		&	Size of image-specific variations	& LogUniform$\left(A_{\rm min}, A_{\rm max}\right)$\\
$\{\ell_1, \ell_2\}$ &	Length scale of image-specific variations & LogUniform$(\ell_{\rm min}, \ell_{\rm max})$\\
$\alpha$	&	Smoothness of image-specific variations	& Uniform$(1, 2)$\\
$f_{\rm bad}$	&	Fraction of ``bad'' data points	& Uniform$(0,1)$\\
$\kappa_1$	&	Error-bar boost for ``good'' data & $\frac{1}{2}\delta(\kappa_1 - 1) + \frac{1}{2}$LogUniform$(1,100)$\\
$\kappa_2$	&	Error-bar boost for ``bad'' data & LogUniform$(\kappa_1, 100\kappa_1)$\\
$\{b_i\}_{i=1}^N$ &	Badness hidden variables for each data point & Uniform$(0,1)$\\
\hline
{\bf Derived Parameter} & {\bf Meaning}	&	{\bf Definition}\\
\hline
$\{B_i\}_{i=1}^N$	&	Badness flag for each data point & $\mathds{1}\left(b_i \leq f_{\rm bad}\right)$\\
\hline
{\bf Data}	&	{\bf Meaning}			& {\bf Sampling Distribution}\\
\hline
$\{Y_1, ..., Y_N\}$ &	Noisy magnitude measurements	& $\mathcal{N}\left(\mathbf{m}(\params), \mathbf{C}(\params)\right)$\\
\hline
{\bf Prior Information} & {\bf Meaning}			&\\
\hline
$N$		&	Number of measurements (all images together)\\
$\{t_1, ..., t_N\}$ &	Measurement times\\
$\{k_1, ..., k_N\}$ &	Flags identifying which image was observed
\end{tabular}
\end{table*}

@@

>>>>

However, any inference of $H_0$ or other cosmological parameters
from the gravitational lensing technique inherits
the uncertainty from the measurements of the time delays themselves. A
thorough investigation of all known sources of uncertainty is an integral part
of measuring cosmological parameters, and thus the purpose of this paper is to
develop and present a reliable method for estimating the uncertainty on the
values of the time delays themselves. The method is Bayesian
\citep{2004kats.book.....O} so the results obtained are probability
distributions over the space of possible solutions for the time delays, and the
uncertainties in the time delays
can be summarised by the widths of these probability distributions. To describe
prior knowledge about the underlying time variation of the quasar and the
effects of microlensing on
the light curves, we use Gaussian Processes (GPs), a family of probability
distributions over function space that has computationally convenient
properties \citep{rasmussen}.

@@

In Section~\ref{sec:bayes} we briefly discuss Bayesian Inference, and in
Section~\ref{sec:sampling} we describe our specific
modelling assumptions. The computational implementation is described in
Section~\ref{sec:computation}. In Section~\ref{sec:tests} we test the inference
with two simulated data sets, one ``easy'' data set and one ``hard'' data set.
In Section~\ref{sec:b1608} we apply the model to B1608+656. Software implementing
the model described in this paper is freely available online at
{\tt www.github.com/eggplantnbren/Flotsam}.

@@

We begin by defining our model for the prior knowledge about how the time
delays give rise to the observed data. This will determine the sampling
distributions, and also the set of parameters (including nuisance parameters) to
be inferred.

\section{Sampling Distributions}\label{sec:sampling}
The intrinsic quasar variability, in magnitudes, and in whatever band the
observations are carried out in, can be described by a function $q(t)$. For a
two-image system, the resulting light curves, if observed without gaps and
noise, would be described by the functions
\begin{eqnarray}
y_1(t) &=& m_1 + q(t) + \mu_1(t) \\
y_2(t) &=& m_2 + q(t - \tau_2) + \mu_2(t) \\
\end{eqnarray}
where $\{m_1, m_2\}$ are the mean magnitudes of the images,
$\tau_2$ is the time delay of image 2 relative to image 1, and the
functions $\{\mu_1(t), \mu_2(t)\}$ describe any additional fluctuations that
are not shared by both images. A major source of additional fluctuations
is the effect of
microlensing on each of the light curves, which appears because the foreground
lens galaxy is actually composed of stars rather than smoothly
distributed matter \citep{falco}.
The data then consist of
noisy, irregularly sampled measurements of the light curves
$\{y_1(t), y_2(t)\}$.

@@

When we use Bayes' rule, the flags $\{k_1, k_2, ..., k_N\}$ and measurement
times $\{t_1, t_2, ..., t_N\}$ will be subsumed into the prior information. The
data (to which we will assign a ``sampling distribution'') will be the
measurements
\begin{eqnarray}
\boldsymbol{D} = \{Y_1, Y_2, ..., Y_N\}.
\end{eqnarray}

The sampling distributions (and hence the likelihood function) will then be
written as a probability distribution for the measurements $\data$
given the time delay $\tau_2$ along with some nuisance parameters. For a full
list of parameters, see Table~\ref{tab:prior}. The total number of parameters
for a two-image system is $13+N$, for a four image system the number of
parameters is $21+N$.

\begin{table*}
\begin{tabular}{lll}
\hline
{\bf Parameter} &	{\bf Meaning}			& {\bf Prior}\\
\hline
$\tau_2$	&	Time delay of image 2	&	Cauchy$\left(0, 0.1\tau_{\rm range}\right)T\left(\tau_{\rm min},\tau_{\rm max}\right)$\\
$\{m_1, m_2\}$	&	Mean magnitude for each image	& Uniform$(m_{\rm min}, m_{\rm max})$\\
$\beta$		&	Size of QSO variations		& LogUniform$(\beta_{\rm min}, \beta_{\rm max})$\\
$L$		&	Length scale of QSO variations	& LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\{S_1, S_2\}$	&	Size of image-specific variations	& LogUniform$\left(A_{\rm min}, A_{\rm max}\right)$\\
$\{\ell_1, \ell_2\}$ &	Length scale of image-specific variations & LogUniform$(\ell_{\rm min}, \ell_{\rm max})$\\
$\alpha$	&	Smoothness of image-specific variations	& Uniform$(1, 2)$\\
$f_{\rm bad}$	&	Fraction of ``bad'' data points	& Uniform$(0,1)$\\
$\kappa_1$	&	Error-bar boost for ``good'' data & $\frac{1}{2}\delta(\kappa_1 - 1) + \frac{1}{2}$LogUniform$(1,100)$\\
$\kappa_2$	&	Error-bar boost for ``bad'' data & LogUniform$(\kappa_1, 100\kappa_1)$\\
$\{b_i\}_{i=1}^N$ &	Badness hidden variables for each data point & Uniform$(0,1)$\\
\hline
{\bf Derived Parameter} & {\bf Meaning}	&	{\bf Definition}\\
\hline
$S_{\rm QSO}$	&	Standard deviation of QSO variability	& $\beta\sqrt{L}$\\
$\{B_i\}_{i=1}^N$	&	Badness flag for each data point & $\mathds{1}\left(b_i \leq f_{\rm bad}\right)$\\
\hline
{\bf Data}	&	{\bf Meaning}			& {\bf Sampling Distribution}\\
\hline
$\{Y_1, ..., Y_N\}$ &	Noisy magnitude measurements	& $\mathcal{N}\left(\mathbf{m}(\params), \mathbf{C}(\params)\right)$\\
\hline
{\bf Prior Information} & {\bf Meaning}			&\\
\hline
$N$		&	Number of measurements (all images together)\\
$\{t_1, ..., t_N\}$ &	Measurement times\\
$\{k_1, ..., k_N\}$ &	Flags identifying which image was observed
\end{tabular}
\caption{A full list of parameters, prior information, and data in the
FLOTSAM model. If there are more than two images, extra parameters are
added accordingly.\label{tab:prior}}
\end{table*}

@@

\bibitem[\protect\citeauthoryear{Falco, Wambsganss, 
\& Schneider}{1991}]{falco} Falco E.~E., Wambsganss J., Schneider P., 1991, MNRAS, 251, 698
